### 2017-02-08 Thesis/latex/paperC.tex
paperC - next step

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Explicit solutions for a long-wave model with constant vorticity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

In nonlinear dispersive evolution equations, the competing effects of nonlinearity and dispersion 
make a number of interesting phenomena possible. In the current work, the focus is on the numerical
approximation of traveling-wave solutions of such equations.
We describe our efforts to write a dedicated \textsf{Python}
code which is able to compute traveling-wave solutions of nonlinear
dispersive equations in a very general form.
%\begin{equation*}
%u_t + [f(u)]_{x} + \Lop u_x = 0,
%\end{equation*}
%%\vskip -0.05cm
%where $\Lop$ is a self-adjoint operator, and $f$ is a real-valued function with 
%$f(0) = 0$ and $f'(0) = 0$.

The \textsf{SpecTraVVave} code uses a continuation method coupled with a spectral projection
to compute approximations of steady symmetric solutions of this equation.
The code is used in a number of situations to gain an
understanding of traveling-wave solutions. The first case is the Whitham
equation, where numerical evidence points to the conclusion that the main 
bifurcation branch features three distinct points of interest, namely a turning point, 
a point of stability inversion, and a terminal point which corresponds to a cusped wave.

The second case is the so-called modified Benjamin--Ono equation where
the interaction of two solitary waves is investigated. It is found
that two solitary waves may interact in such a way
that the smaller wave is annihilated. The third case concerns
the Benjamin equation which features two competing dispersive operators.
In this case, it is found that bifurcation curves of periodic traveling-wave
solutions may cross and connect high up on the branch in the nonlinear regime.
%Include keywords and mathematical subject classification numbers as needed.

\subsection{Introduction}
\label{intro}

This paper is concerned with traveling wave solutions for a class of nonlinear dispersive
equations of the form
\begin{equation}\label{nonlocal}
u_t + [f(u)]_x + \Lop u_x = 0,
\end{equation}
where $\Lop$ is a self-adjoint operator, and $f$ is a real-valued function 
with $f(0)=0$ and $f'(0)=0$, and which satisfies certain growth conditions.
Equations of this form arise routinely in the study of wave problems in fluid mechanics
and many other contexts.
A prototype of such an equation is the KdV equation that appears
if $\Lop= I + \frac{1}{6} \partial_x^2$ and $f(u) = \frac{3}{4} u^2$.
In the current work, the operator $\Lop$ is considered to be given as a Fourier multiplier operator,
such as for instance in the Benjamin--Ono equation, which arises in the study of
interfacial waves. In this case, the Fourier multiplier operator is given by
$\Lop =  I - \Hop \partial_x$, where the Hilbert transform
$\Hop$ is defined as
\begin{equation}
\Hop u(x) = \frac{1}{\pi} \, \mathrm{p.v.} \int_{-\infty}^{\infty} \frac{u(x-y)}{y} \, \dd y, 
\qquad \widehat{\Hop u}(k) = -\ii \operatorname{sgn}(k)\widehat{u}(k).
\end{equation}
We also study in detail traveling wave solutions of the Whitham equation, which appears when 
$\Lop$ is given by convolution with the integral kernel $K_{h_0}$ in the form
\begin{equation}\label{symbol}
\Lop u(x) = \int_{-\infty}^{\infty} K_{h_0} (y) u(x-y) \, \dd y,
\qquad \widehat{K_{h_0}} (k) 
= {\textstyle{ \sqrt{\frac{g\tanh (h_0 k)}{k}}}}, 
\end{equation}
and $f$ is the same function as in the KdV equation.

The particular form of equation~\eqref{nonlocal} exhibits the competing effects of dispersion and nonlinearity, 
which gives rise to a host of interesting phenomena. The most well known special phenomenon is the existence
of solitary waves and of periodic traveling waves containing higher Fourier modes.
Indeed, note that in the purely dispersive model $u_t + \Lop u_x = 0$, 
the only possible permanent progressive waves are simple sinusoidal waves, while
in the nonlinear model~\eqref{nonlocal} higher Fourier modes must be considered
to obtain solutions.

The order of the operator $\mathcal{L}$ appearing in~\eqref{nonlocal} 
has a major effect on the types of solutions that may be found. 
A higher-order operator, such as in the Korteweg--de Vries equation, 
acts as a smoothing operator because of its effect of spreading different frequency components
out due to a strongly varying phase speed~\cite{Kato1983}. Lower-order operators 
such as the operator $K_{h_0}$ in~\eqref{symbol} appearing in the Whitham equation
may allow solutions to develop singularities, such as derivative blow-up (see~\cite{Hur2015arxiv}, \cite{Tao2014})
and formation of cusps (see~\cite{Wahlen2016}).

On the other hand, highly nonlinear functions $f(u)$ may lead to $L^{\infty}$-blow-up.
For instance, the generalized KdV equation which is written in normalized form as
\begin{equation}
\label{kdv}
u_t + u^{p}u_x + u_x + u_{xxx} = 0,
\end{equation}
features global existence of solutions for $p=1,2,3$, 
but the solutions blow-up in the critical case $p=4$ (the case $p \ge 5$ is open).
In the case of the generalized Benjamin--Ono equation
\begin{equation*}
u_t + u^{p}u_x + u_x - \mathcal{H} u_{xx} = 0,
\end{equation*}
where $\mathcal{H}$ is the Hilbert transform, numerical evidence points to singularity formation 
for $p > 2$~\cite{Bona2004}, but no proofs are available at this time.

In order to study different phenomena related to equations of the form~\eqref{nonlocal} 
and their traveling wave solutions, a \textsf{Python}-based solver package \textsf{SpecTraVVave} 
was developed by the authors~\cite{github}. 
The general idea behind the solver is to use a numerical continuation method~\cite{Keller1977} 
implemented with a pseudo-spectral algorithm. 
Similar previous projects include \textsf{AUTO}~\cite{Doedel1997} and \textsf{Wavetrain}~\cite{Sherratt2012}.
\textsf{AUTO} is written in \emph{C}, whereas \textsf{Wavetrain} is written in \emph{Fortran}. 
Both programs are efficient and very general, as they are able to cover a wide range of problems
involving bifurcation analyses. However, from a user's perspective, 
such a generality coupled with low level programming languages
may lead to some difficulties in utilizing these programs efficiently.

\textsf{SpecTraVVave} is designed to provide researchers with a simple yet effective tool 
for investigating problems on traveling waves. The package is flexible, 
and its functionality can be easily expanded. 
The availability of the \textsf{IPython} notebook \cite{Perez2007} makes the solver very interactive,
so that it should be easier for new users to get started.

In order to maximize ease of use, \textsf{SpecTraVVave} was designed
to find even solutions of~\eqref{nonlocal}.
Symmetry of steady solutions can be proved for some of the models
in the form~\eqref{nonlocal}, but not for all~\cite{Chen1998}.
Some of these equations also admit non-smooth solutions,
for instance as termination points of a bifurcation branch.
This happens for example for the Whitham equation, which features bifurcation
curves which terminate in a solution with a cusp~\cite{Wahlen2016}.
One of the goals of the present paper is to investigate the
precise nature of the termination of the bifurcation curve.

The content of the paper is structured as follows. A mathematical description of the numerical method of 
\textsf{SpecTraVVave} is given in \autoref{sec:spectral}. \autoref{sec:experiments} presents results of different experiments carried out with the package. Concluding remarks are given in \autoref{sec:conclusion}. A method for finding initial guesses for the solver is described in \autoref{sec:stokes}. \autoref{sec:implementation} contains a schematic of program and a description of its workflow.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spectral scheme and construction of nonlinear system.}
\label{sec:spectral}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Cosine collocation method.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
To compute traveling wave solutions to the equation~\eqref{nonlocal} the following ansatz is employed:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation*}%\label{ansatz}
u(x,t) = \phi(x-ct).
\end{equation*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Thus, the equation takes the form
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation*}%\label{phi-nonlocal}
\phi' + \left[f(\phi) \right]' + \Lop \phi' = 0,
\end{equation*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
which can be integrated to give
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}\label{int_gen1}
-c \phi + f(\phi)  + \Lop \phi = B. 
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The constant $B$ is a priori undetermined. One may set the $B$ equal to 
zero as a way of normalizing the solutions. Another option is to impose
an additional condition, for example that the integral of $\phi$ over
one wavelength be zero. In this case, $B$ will be found along with the solution $\phi$.

We consider $\Lop$ as a Fourier multiplier operator with symbol
$\alpha(k)$. We also assume that $f$ is at least twice differentiable,
and we have $f(0)=0$ and $f'(0)=0$.
When computing traveling-wave solutions we focus on even periodic solutions. 
While it can be proved in some cases that solutions of~\eqref{int_gen1}
must be even, this is not known for a general operator $\Lop$.
Nevertheless, we make this assumption here in order to
make the numerical procedure as uniform as possible.
For even periodic solutions, 
one may use a cosine collocation instead of a Fourier method. In particular, using
the cosine functions as basis elements automatically removes the inherent symmetries
due to reflective and translational symmetry. 
Moreover, the number of unknowns is reduced by a factor of $2$, 
and the problem of the asymmetric arrangement of nodes in the FFT
is circumvented. Of course, all these problems could also be dealt with a collocation
method based on the Fourier basis, but the cosine basis does all of the above automatically.
In addition,  the \textsf{Python} cosine transform is based on an integrated algorithm, which
relies on an optimized version of the discrete cosine transform (DCT).

The following description of computation scheme was presented in detail in~\cite{Ehrnstrom2013}, 
but we will briefly repeat it here for consistency of the manuscript. For the purpose of clarity, 
we will refer to full wavelength $L$ of a solution as the (full) wavelength, 
and half of fundamental wavelength will be called half-wavelength. 
Such a definition is required because the method computes a half of a solution profile, 
the other half is automatically constructed due to symmetry.

Traveling wave solutions to the equation~\eqref{int_gen1}
are to be computed in the form of a linear combination of cosine functions of different
wave-numbers, i.e., in the space
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{equation}
\SN = \mathrm{span}_{\R}\setc{\cos(lx)}{0 \leq l \leq N-1 }
.
\end{equation}
%%%%%%%%%===========%%%%%%%%%%%%%%%
This is a subspace of $L^2(0,2\pi)$, and the collocation points 
$x_n = \pi \frac{2n -1}{2N}$ for $n = 1,\dots, N$ are used to discretize the domain.  
If the required full wavelength of solutions is to be $L \neq 2\pi$, 
one can use a scaling on the $x$-variable. 
Defining the new variable
\begin{equation*}
x' = \frac{L}{2\pi}x, 
\end{equation*}
yields collocation points $x_n'$ and wavenumbers $\kappa_l$ defined by
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{equation*}
x_n' = \frac{L}{2} \frac{2n -1}{2N}, \quad\quad \kappa_l = \frac{2\pi}{L}l. 
\end{equation*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
%%%%%%%%%===========%%%%%%%%%%%%%%%
%%%%%%%%%===========%%%%%%%%%%%%%%%
We are seeking a function $\phi_N \in \SN$ that satisfies the equations
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{equation}
-c\phi_N(x_n') + f(\phi_N)(x_n') + \Lop^N\phi_N(x_n') = 0, \label{int_gen-disc}
\end{equation}
%%%%%%%%%===========%%%%%%%%%%%%%%%
at the collocation points $x_n'$.
The operator $\Lop^N$ is the discrete form of the operator $\Lop$, 
and $\phi_N$ is the discrete cosine representation of $\phi$
which is given by
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align*}
\phi_N(x') &= \sum_{l = 0}^{N-1} \omega(\kappa_l)\Phi_N(\kappa_l)\cos(\kappa_{l}x'),\\
\omega(\kappa_l) &= 
\begin{cases}
\sqrt{1/N}, \quad \kappa_l = 0, \\
\sqrt{2/N}, \quad \kappa_l > 0,
\end{cases}                  \\
\Phi_N(\kappa_l) &= \omega(\kappa_l) \sum_{n = 1}^{N} \phi_N(x'_n) \cos(\kappa_l x'_n),
\end{align*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
where $\kappa_l = 0,\frac{2\pi}{L},\ldots, \frac{2\pi}{L}(N-1)$ are the scaled wavenumbers,
and $\Phi_N(\cdot)$ are the discrete cosine coefficients.
As the equation~\eqref{int_gen-disc} is enforced at the collocation points $x_n'$, 
one may evaluate the term $\Lop^N\phi_N$ using the matrix $\Lop^N(i,j)$ 
defined by
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align*}
\Lop^N\phi_N(x'_i) &= \sum_{j=1}^{N}\Lop^N(i,j) \phi_N(x'_j),\\
\Lop^N(i,j) &= \sum_{l=0}^{N-1} \omega^2(\kappa_l) \alpha(\kappa_l) \cos(\kappa_l x'_i) \cos(\kappa_l x'_j),
\end{align*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
where $\alpha(\cdot)$ is the Fourier multiplier function of the operator $\Lop$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Construction of nonlinear system.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
The equation~\eqref{int_gen-disc} enforced at $N$ collocation points 
yields a nonlinear system of $N$ equations in $N$ unknowns, which can be written in shorthand as  
%%%%%%%%%%===========%%%%%%%%%%%%%%%
	\begin{equation*}
		F(\phi_N) = 0.
	\end{equation*}
%%%%%%%%%%===========%%%%%%%%%%%%%%%
This system can be solved by a standard iterative method, such as Newton's method. 
%%%%%%%%%%===========%%%%%%%%%%%%%%%
	In this system, the value of phase speed $c$ has to be fixed for computing one particular solution. 
	Such an approach becomes impractical when a turning point on the bifurcation curve appears.

	In \textsf{SpecTraVVave} a different approach is employed: both the amplitude $a$ and the phase speed $c$ of a solution are treated as functions of a parameter $\theta$: $ a = a(\theta)$, $c = c(\theta)$. The parameter $\theta$ is to be computed from the system~\eqref{system-N3}. This construction makes it possible to follow turning points
on the bifurcation branch with relative ease.
	Having computed two solutions, i.e., two points on the bifurcation curve $P_1 = (c_1, a_1)$ and $P_2 = (c_2, a_2)$, one may find a direction vector $\mathbf{d}=(d^c, d^a)$ of the line that contains these points:
%%%%%%%%%===========%%%%%%%%%%%%%%%
	\begin{align*}
		\mathbf{d}:	\quad d^c = c_2 - c_1, \qquad	d^a = a_2-a_1.
	\end{align*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
	Then the point $P_3 = (c_3,a_3)$ is fixed at some (small) distance $s$ from the point $P_2$ in the direction $\mathbf{d}$.
%%%%%%%%%===========%%%%%%%%%%%%%%%
	\begin{align*}
		P_3: \quad c_3 = c_2 + s \cdot d^c, \qquad	a_3 = a_2 + s \cdot d^a.
	\end{align*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
	 The point $P_3$ plays the role of the initial guess for velocity and amplitude when computing the next solution $P_\ast = (c_\ast, a_\ast)$. 
	The solution point $P_\ast$ is required to lay on the line with direction vector $\mathbf{d_{\bot}}=(d^c_{\bot}, d^a_{\bot})$, which is orthogonal to the vector $\mathbf{d}$.
%%%%%%%%%===========%%%%%%%%%%%%%%%
	\begin{align*}
	\mathbf{d_{\bot}}: \quad d^c_{\bot} = - d^a, \qquad &d^a_{\bot} = d^c,  \\
	P_\ast: \quad c_\ast = c_3 + \theta d^c_{\bot} \qquad &a_\ast = a_3 + \theta d^a_{\bot}.
	\end{align*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
	 A schematic sketch of finding a new solution $P_\ast$ is given in Figure 1.

	The variable $\theta$ is computed by Newton's method from the extended system 
%%%%%%%%%===========%%%%%%%%%%%%%%%	
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{equation}
F\left(
\begin{matrix}
\phi_N(x_1)\\ \vdots \\ \phi_N(x_N) \\ B \\ \theta
\end{matrix}
 \right)= \left(
\begin{matrix}
(-c+\mathcal{L}_N)\phi_N(x_1)+ f(\phi_N(x_1)) - B
\\ \vdots 
\\ (-c+\mathcal{L}_N)\phi_N(x_N)+ f(\phi_N(x_N)) - B
\\ \Omega(\phi_N, c, a, B)
\\ \phi_N(x_1)-\phi_N(x_N)-a
\end{matrix}
 \right)=
\left(
\begin{matrix}
0
\\ \vdots 
\\ 0
\\ 0
\\ 0
\end{matrix}
 \right). \label{system-N3} 
\end{equation}
%%%%%%%%%===========%%%%%%%%%%%%%%%
	Here a nonhomogeneous problem ($B \neq 0$) is considered.
	The equation 	
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{equation*}
	\phi_N(x_1)-\phi_N(x_N)-a = 0, 
\end{equation*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
	 makes the waveheight of the computed solution to be that of $a$. The equation 
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{equation*}
	\Omega(\phi_N, c, a, B)=0, %\label{boundary}
\end{equation*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
	 is called the  \emph{boundary condition}. 
	It allows to enforce different specifications on the computed traveling wave solution.   	
	For example, if one sets  
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{equation*}
	\Omega(\phi_N, c, a, B)= \phi_N(x_1)+\cdots+\phi_N(x_N), 
\end{equation*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
	 then \emph{the mean of the computed wave over a period will have to be equal to zero}.
	One may also experiment with	
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{equation*}
	\Omega(\phi_N, c, a, B) = B, 
\end{equation*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
	 to consider the homogeneous problem $(B=0)$.
	It can be also interesting to set
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{equation}
	\Omega(\phi_N, c, a, B) = \phi_N(x_N). \label{solitary}
\end{equation}
%%%%%%%%%===========%%%%%%%%%%%%%%%	
	 This enables us to compute traveling wave solutions that mimic solitary wave solutions. 	
%%%%%%%%%===========%%%%%%%%%%%%%%%
%\begin{figure}
%\centering
%		\includegraphics[width=0.35\textwidth]{navig}
%\caption{\small Navigation on the bifurcation curve.}
%\label{navig}
%\end{figure} 
%%%%%%%%%===========%%%%%%%%%%%%%%%

\subsection{Convergence.}
In order to test the numerical implementation of the discretization,
the method is applied to a case where the solution is known.
One such case is the KdV equation
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align*}
u_t + u_x + \frac{3}{2} u u_x + \frac{1}{6}u_{xxx} = 0, 
%\label{KdV_sol*}
\end{align*}
which has a known solution, given in the form
\begin{align*}
u_{\mathrm{exact}}(x,t) = a \sech^2\Big( {\textstyle\sqrt{\frac{3a}{4}}} (x-c t)\Big),\\
\end{align*}
with $c=1+a/2$.
%%%%%%%%%===========%%%%%%%%%%%%%%%
%
Using the boundary equation~\eqref{solitary}, \textsf{SpecTraVVave} is capable of computing 
approximations to solitary wave solutions of nonlinear wave equations. 
Solitary wave solutions are treated as traveling waves with sufficiently long wavelength 
that have the wave trough at zero.
In case of the KdV equation solitary wave solutions have exponential decay, 
and therefore, considering the symmetry of solitary solutions, 
the half-wavelength of $30$ is considered for the comparison. 
%\itodo[DM]{as mentioned before: should we call the (full-wavelength )/2 as half-wavelength OR just wavelength. 
%This may lead to confusions for reviewers.}
Approximation errors are summarized in \autoref{t2}.
%
\begin{table}  [ht]
\centering
\begin{tabular}{rrrr}
  \toprule
Nb.\ of grid points & $\log_{10}(\|u_{\mathrm{exact}} - u\|_{L^{\infty}})$ & $\log_{10}(\|u_{\mathrm{exact}} - u\|_{L^2})$ & Ratio of $L^2$-errors\\
\midrule
32 & $-1.389$  & $-2.092$ & \\

64 & $-3.705$  & $-4.549$ & $286.8$\\

128 & $-8.809$  & $-9.508$ & $90935.0$\\

256 & $-15.353$  & $-16.144$ & $4329670.9$\\

512 & $-15.353$  & $-16.087$ & $0.9$\\
\bottomrule
\end{tabular}
\caption{Estimates of error between the exact and computed solitary wave solutions for the KdV equation. Half-wavelength $30$, waveheight $a=1.2651$}
\label{t2}
\end{table}
%%%%%%%%%===========%%%%%%%%%%%%%%%

%%%%%%%%%===========%%%%%%%%%%%%%%%
%\begin{figure}[ht]
%\centering
%\includegraphics[width=0.47\textwidth]{convergence}      
%\caption{\small Graph of error estimates given in \autoref{t2}.}
%\label{L2convergence}
%\end{figure} 
%%%%%%%%%===========%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiments with SpecTraVVave.}
\label{sec:experiments}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsubsection{Termination of the waveheight-velocity bifurcation curve of the Whitham equation.}
\label{sec-W-term}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The waveheight-velocity bifurcation curve of the Whitham equation 
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align}
	u_t + \frac{3}{2} u u_x + Ku_x = 0, \qquad 
	\widehat{Ku}(k) = \sqrt{\frac{\tanh(k)}{k}},  \label{W}
\end{align}
%%%%%%%%%===========%%%%%%%%%%%%%%%
was studied numerically in~\cite{Ehrnstrom2013}. 	
An attempt was made to identify the termination point of the Whitham bifurcation curve. 
The investigation was limited by computational tools and complete results were not obtained. 
In particular, the authors could not confirm that traveling wave solutions do not exist 
past the point where the authors, based on pioneering work of Whitham
\cite{Whitham1967j} suspected a cusped solution.
In this section a number of numerical results on nature of the bifurcation curve
for the Whitham equation are presented.
Solutions to the equation~\eqref{W}
are computed in the form of traveling waves $u(x,t) = \phi(x - c t)$
and the homogeneous $(B = 0)$ integrated version the equation is considered:
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align}
-c \phi + \frac{3}{4} \phi^2 + K \phi = 0. \label{W-int}
\end{align}
%%%%%%%%%===========%%%%%%%%%%%%%%%
Special attention is given to relation between stability of solutions 
and their waveheight and velocity parameters, i.e., their position on the bifurcation curve. 
The following questions are under study: 
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{enumerate}[label=\alph*)] % chktex 9 chktex 10
\item Where does the bifurcation curve terminate?

\item Where on the bifurcation curve do solutions change their stability?

\item Is there any role that the turning point on the bifurcation curve plays?
\end{enumerate}
%%%%%%%%%===========%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\subsubsection{Implementation of the problem in SpecTraVVave.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The results presented here focus on $2\pi$-periodic solutions to the equation~\eqref{W-int}, 
i.e., solutions of the system~\eqref{system-N3}. 
Figure 3 presents Whitham bifurcation curves with numbers of grid points 
$N = 512$, $N=1024$ and $N=2048$. 
The current implementation of the \textsf{SpecTraVVave} package allows fixing the number of grid points $N$ 
and a so-called doubling parameter $\DD$, 
i.e., the number by which $N$ is doubled as computations are made. 	
This allows us to get sets of solutions with $N, 2N, \ldots, 2^{\DD} N$ grid points. 
If $\DD = 1$ then only two sets of solutions are computed and they are regarded as 
lower grid (lower resolution) and higher grid (higher resolution) solutions.
While the system~\eqref{system-N3} is processed by Newton solver, 
lower grid solutions are taken as initial guesses for higher grid solutions. 
All curves shown in this manuscript have been produced
after tests with a number of resolutions were run, 
and the curves shown did not change significantly under further refinement.

%%%%%%%%%===========%%%%%%%%%%%%%%%
%\begin{figure}[ht]
%\centering
%\includegraphics[width=0.48\textwidth]{512_1024_2048}{}      
%\caption{\small Whitham bifurcation curve in different grid resolutions.}
%\label{bif-curve-all}
%\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%\subsubsection{Specific solutions and their stability.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
Figure 4(a) % chktex 8
presents the Whitham bifurcation curve computed by \textsf{SpecTraVVave} 
with $N=1024$ and $\DD=1$. 
There are three solutions which deserve to be singled out:
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{enumerate}
	\item Traveling wave solution with minimum velocity (rhombus);
	\item Traveling wave solution with maximum $L^2$-norm (circle);
	\item Cusped traveling wave solution (square).
\end{enumerate}
%%%%%%%%%===========%%%%%%%%%%%%%%%
Profiles of the above listed solutions are given in Figure 5(a).
The solution with minimum velocity corresponds to the turning point of the bifurcation curve.
The solution with maximum $L^2$-norm is very close to the latter one, although it has a higher waveheight 
and a different velocity.
The solution marked by a square is called here the \emph{terminal solution}.
As already mentioned, previous studies, such as~\cite{Ehrnstrom2009,Ehrnstrom2013} did not provide any conclusive
analysis on the part of the bifurcation curve past the turning point.
In particular, it was not clear whether solutions ceased to exist
at or after the turning point, or whether solutions were stable
or unstable after the turning point.

Let us first focus on the stability of solutions.
Note that \textsf{SpecTraVVave} has an evolution integrator routine, 
which enables one to check the stability of computed solutions. 	
The current version of the package uses the fourth-order method developed in~\cite{DeFrutos1992}.
In addition one may use a more refined analysis, resting on the evaluation
of invariant functionals. This analysis is based on the observation
that the traveling waves can be thought of as solutions of a constrained
minimization problem. This analysis is based on ideas developed by
Boussinesq, first exploited in~\cite{Benjamin1972j}, and later used 
in~\cite{Bona1987,Bridges2004,Nguyen2009}, and many other works.
%%%%%%%%%===========%%%%%%%%%%%%%%%
%\begin{figure}[ht]
%\centering
%\subfigure[][Bifurcation curve in the wavespeed-waveheight parameter space]{\includegraphics[width=0.48\textwidth]{1024_2048}{}\label{sfig:bif-ch}}%
%~%
%\subfigure[][Bifurcation curve in a wavespeed-$L^2$-norm diagram]{\includegraphics[width=0.48\textwidth]{L2_1024_2048}{}} 
%\caption{\small
%Whitham bifurcation curves for $2\pi$-periodic solutions. 
%% $N = 1024, \mathcal{D}= 1$. 
%}
%\label{bif-L2-1024}
%\end{figure} 
%trim = 1.2cm 4.7cm 0.6cm 4.5cm, clip,
%%%%%%%%%===========%%%%%%%%%%%%%%%
Let us define two functionals $V$ and $E$:
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align*}
	V( \phi) = \frac{1}{2} \int_{-\infty}^{+\infty} \phi^2\,\dd\zeta, \qquad
	E( \phi) = \int_{-\infty}^{+\infty} \left\{ \Sfrac{1}{2} \phi^3 - \phi \, K\!  \phi \right\}\, \dd\zeta.
\end{align*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
The equation~\eqref{W-int} can be then written in terms of variational derivatives of $E$ and $V$
as
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align}
	E' ( \phi ) - c V ' ( \phi ) = 0.
\label{zero}
\end{align} 
%%%%%%%%%===========%%%%%%%%%%%%%%%
It is known from~\cite{Bona1987} that under certain conditions,
the stability of solitary wave solutions depends 
on convexity of the function $d(c) = E(\phi) - c V(\phi)$. 
Solutions with values of $c$ for which $d''(c) > 0$  are stable solutions,
and solutions with wave speeds for which $d''(c) <0$ are unstable solutions.

The current numerical investigation can therefore be interpreted as an indication of the
stability properties of traveling wave solutions of the Whitham equation.
Note that differentiation of $d(c)$ yields
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align*}
	d '(c) &= \underbrace{E'(\phi) - c V'(\phi)}_{=0} - V(\phi).
\end{align*}
Using~\eqref{zero} as indicated yields
\begin{align}
	d '(c) &= - V(\phi) = - \frac{1}{2} \int_{-\infty}^{+\infty} \phi ^2\,\dd\zeta = - \frac{1}{2}\|\phi\|_{L^2}^2 . 
\label{d'c}
\end{align}
%%%%%%%%%===========%%%%%%%%%%%%%%%
Therefore, in order to understand the convexity of $d(c)$, it is sufficient
to find points of maximum $L^2$-norm on the curve in the right panel
of Figure 4.
It is straightforward to see that $d''(c)$ changes sign in the neighbourhood of the maximum point 
of this curve, i.e., around the solution with maximum $L^2$-norm. 
In particular, $d''(c) > 0$, i.e., solutions are stable to the left of the maximum point, 
and $d''(c) < 0$, i.e., solutions are unstable to the right of the maximum point.

In addition, the solutions were tested with the evolution integrator to confirm their stability/instability in time. 
The solution with maximum $L^2$-norm and those on the left-hand side were always found to be stable in 
the time-dependent computations.
Solutions on the right-hand side do not preserve their shape and thus are unstable. 
Examples are given in Figure 6. 
This analysis confirms that 
the point corresponding to the minumum wave speed (the turning point), 
and the point of stability inversion are two distinct points on the bifurcation curve.
Moreover, the point of stability inversion
is a little further up the branch from the turning point.

%%%%%%%%%===========%%%%%%%%%%%%%%%
%\begin{figure}[t]
%\centering
%\subfigure[][Profiles of the solutions singled out in \autoref{bif-L2-1024}\subref{sfig:bif-ch}]{\includegraphics[width=0.48\textwidth]{1024_waves_1}{}\label{sfig:prof-sing}}
%~%
%\subfigure[][Profile after terminal solution]{\includegraphics[width=0.48\textwidth]{1024_waves_2}{}\label{sfig:prof-term}}
%\caption{\small Profiles of specific traveling wave solutions.}
%\label{3waves}
%\end{figure} 
%%%%%%%%%===========%%%%%%%%%%%%%%%

Next, we turn our attention to the analysis of the terminal point. There
are two main questions. Does the branch terminate, and if so, 
does the terminal point on the branch correspond to a cusped traveling wave.
First of all, note that the solution, which is computed by \textsf{SpecTraVVave}, 
past the terminal solution has two crests, no matter how small the 
stepping on the bifurcation branch is taken.
(see Figure 5(b)).   
Secondly, as will be explained presently, the relation 
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{equation*}
\frac{c}{\sup_{x} \phi(x)} = \frac{3}{2} 
\end{equation*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
holds for the terminal solution with a good degree of approximation.
For the most accurate runs, we obtain $c / \sup_{x \in \R} \phi(x)$ $\approx$ $1.51$. 
%
To explain how this relation comes about note that the steady integrated form of the
Whitham equation can be written as
\begin{equation}\label{Whitham_Alt}
\Big( \frac{c}{\sqrt{3}} - \frac{\sqrt{3}}{2}\phi \Big)^{2}  = \frac{1}{3} c^2 - K \phi.
\end{equation}
It is clear that for any $\phi < \frac{3}{2} c$, the relation~\eqref{Whitham_Alt}
can used in a bootstrap argument to show that any continuous solution must be
in fact smooth. However for the case $\phi =  \frac{3}{2} c$ this bootstrap
argument fails since the left-hand side vanishes. It can be concluded that
a solutions containing a cusp will have a maximum  value of $\frac{3}{2} c$.
